Reanalysis Step 2: Multiple Imputation

### 2025-08-29 reanalysis-2-multiple_imputation.qmd
bugfix

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 30 imputations, 20 iterations (stability achieved ~20)
analysis_mode <- "final"  # Change to "final" for production run

@@

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr"  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr
)

@@

@@

# 2. Intermediate can predict outcome
pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1

@@

@@

```{r mice-run}

@@

  # Create summary table for cataract surgery proportions
  if(is.matrix(cat_props)) {
    cat_summary <- data.frame(
      Level = rownames(cat_props),
      Mean_Proportion = apply(cat_props, 1, mean) |> round(3),
      SD = apply(cat_props, 1, sd) |> round(3)
    )
  } else {
    cat_summary <- data.frame(
      Level = "Cataract Surgery",
      Mean_Proportion = round(mean(cat_props), 3),
      SD = round(sd(cat_props), 3)
    )
  }
  kable(cat_summary, caption = "Cataract surgery proportions across imputations")
}
```

>>>>

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 30 imputations, 20 iterations (stability achieved ~20)
analysis_mode <- "test"  # Change to "final" for production run

@@

## Create Fellow Eye Variables

```{r create-fellow-eye-vars}
# Identify patients with both eyes and create fellow eye predictors
# This leverages the correlation between eyes from the same patient
# to improve imputation plausibility

df_with_fellow <- df_for_imputation |>
  group_by(e_mrn_deidentified) |>
  mutate(
    n_eyes = n(),
    has_fellow = n_eyes == 2,
    eye_index = row_number(),

    # Get fellow eye values (NA if patient has only 1 eye)
    # Using [3 - eye_index] to get the other eye (works for 2 eyes: eye 1 gets eye 2, eye 2 gets eye 1)
    fellow_baseline_VA = if_else(n_eyes == 2, 
                                 baseline_VA_logMAR[3 - eye_index], 
                                 NA_real_),
    fellow_intermediate_VA = if_else(n_eyes == 2,
                                    intermediate_VA_logMAR[3 - eye_index],
                                    NA_real_),
    fellow_outcome_VA = if_else(n_eyes == 2,
                               outcome_VA_logMAR[3 - eye_index],
                               NA_real_),
    fellow_lapsed = if_else(n_eyes == 2,
                           ever_lapse_binary[3 - eye_index],
                           NA_real_),
    # Note: fellow_dr needs special handling to maintain factor structure
    fellow_dr_numeric = if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               NA_real_),

    # Patient-level summaries (same for both eyes of a patient)
    patient_worst_baseline = max(baseline_VA_logMAR, na.rm = TRUE),
    patient_best_baseline = min(baseline_VA_logMAR, na.rm = TRUE),
    patient_ever_lapsed = max(ever_lapse_binary, na.rm = TRUE)
  ) |>
  ungroup() |>
  mutate(
    # Handle Inf values from max/min with all NAs
    patient_worst_baseline = if_else(is.infinite(patient_worst_baseline), NA_real_, patient_worst_baseline),
    patient_best_baseline = if_else(is.infinite(patient_best_baseline), NA_real_, patient_best_baseline),

    # Convert fellow_dr_numeric to ordered factor matching person_dr structure
    fellow_dr = factor(fellow_dr_numeric, levels = c(0, 1, 2),
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE)
  ) |>
  select(-fellow_dr_numeric)  # Remove the temporary numeric version

# Summary of eye pairing
n_patients <- n_distinct(df_with_fellow$e_mrn_deidentified)
n_single_eyes <- sum(df_with_fellow$n_eyes == 1)
n_paired_eyes <- sum(df_with_fellow$n_eyes == 2)

# Log patient/eye summary
log_message("Patient/Eye Summary:")
log_message(sprintf("- Total patients: %d", n_patients))
log_message(sprintf("- Patients with 1 eye: %d (%.1f%%)", 
            n_single_eyes, 100 * n_single_eyes / nrow(df_with_fellow)))
log_message(sprintf("- Patients with 2 eyes: %d (%.1f%%)", 
            n_paired_eyes/2, 100 * n_paired_eyes / nrow(df_with_fellow)))
log_message(sprintf("- Total eyes: %d", nrow(df_with_fellow)))

# Display summary for output
patient_eye_summary <- tibble(
  Category = c("Total patients", "Patients with 1 eye", "Patients with 2 eyes", "Total eyes"),
  N = c(n_patients, n_single_eyes, n_paired_eyes/2, nrow(df_with_fellow)),
  Percent = c(NA, 100 * n_single_eyes / nrow(df_with_fellow), 
              100 * n_paired_eyes / nrow(df_with_fellow), NA)
)
kable(patient_eye_summary, caption = "Patient/Eye Distribution", digits = 1)

# Check fellow eye data availability
fellow_summary <- df_with_fellow |>
  filter(n_eyes == 2) |>
  summarise(
    n_fellow_baseline = sum(!is.na(fellow_baseline_VA)),
    n_fellow_intermediate = sum(!is.na(fellow_intermediate_VA)),
    n_fellow_outcome = sum(!is.na(fellow_outcome_VA)),
    pct_fellow_baseline = round(100 * n_fellow_baseline / n(), 1),
    pct_fellow_intermediate = round(100 * n_fellow_intermediate / n(), 1),
    pct_fellow_outcome = round(100 * n_fellow_outcome / n(), 1)
  )

# Log fellow eye data availability
log_message("Among paired eyes, fellow eye data availability:")
log_message(sprintf("- Fellow baseline VA: %.1f%%", fellow_summary$pct_fellow_baseline))
log_message(sprintf("- Fellow intermediate VA: %.1f%%", fellow_summary$pct_fellow_intermediate))
log_message(sprintf("- Fellow outcome VA: %.1f%%", fellow_summary$pct_fellow_outcome))

# Display as table
fellow_availability <- tibble(
  Measure = c("Fellow baseline VA", "Fellow intermediate VA", "Fellow outcome VA"),
  Percent_Available = c(fellow_summary$pct_fellow_baseline, 
                        fellow_summary$pct_fellow_intermediate, 
                        fellow_summary$pct_fellow_outcome)
)
kable(fellow_availability, caption = "Fellow Eye Data Availability (Among Paired Eyes)", digits = 1)

# Use this enhanced dataset for imputation
df_for_imputation <- df_with_fellow
```

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr",  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr

  # Fellow eye predictors (if available)
  "fellow_baseline_VA",
  "fellow_intermediate_VA",
  "fellow_outcome_VA",
  "fellow_lapsed",
  "fellow_dr",
  "n_eyes",
  "has_fellow",
  "patient_worst_baseline",
  "patient_best_baseline",
  "patient_ever_lapsed"
)

@@

# Fellow eye variables should not be imputed (they are derived from existing data)
# NOTE: fellow_outcome_VA may show collinearity warning with outcome_VA_logMAR
# This is expected - they have identical missingness patterns (both NA for excluded cohort)
custom_methods["fellow_outcome_VA"] <- ""
custom_methods["fellow_baseline_VA"] <- ""
custom_methods["fellow_intermediate_VA"] <- ""
# custom_methods["fellow_lapsed"] <- ""  # REMOVED - collinear with lapse status
custom_methods["fellow_dr"] <- ""
custom_methods["n_eyes"] <- ""
# custom_methods["has_fellow"] <- ""  # REMOVED - collinear with n_eyes
custom_methods["patient_worst_baseline"] <- ""
custom_methods["patient_best_baseline"] <- ""
# custom_methods["patient_ever_lapsed"] <- ""  # REMOVED - collinear with ever_lapse_binary

@@

# 2. Intermediate can predict outcome
# NOTE: MICE predictor matrix only uses 0/1, not weights.
# To prioritize intermediate VA over baseline for outcome prediction:
# - Both are included as predictors (temporal proximity naturally helps)
# - PMM will naturally weight closer timepoints more heavily
# - Alternative: Could create interaction terms or use method="norm.boot" with custom weights
pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1

@@

# DR severity predicts visual acuity at all timepoints
if ("person_dr" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "person_dr"] <- 1
  pred_matrix["intermediate_VA_logMAR", "person_dr"] <- 1
  pred_matrix["outcome_VA_logMAR", "person_dr"] <- 1
}

# FELLOW EYE STRUCTURE
# Fellow eye outcomes strongly predict index eye outcomes (correlation typically 0.7-0.9)
# NOTE: MICE predictor matrix only uses 0/1, not weights. 1 = used as predictor, 0 = not used
if ("fellow_outcome_VA" %in% colnames(pred_matrix)) {
  # Fellow eye visual acuity predicts same timepoint
  pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
  pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1  
  pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 1

  # Fellow baseline also helps predict later timepoints
  pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
  pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1

  # DR severity correlation between eyes
  if ("person_dr" %in% rownames(pred_matrix)) {
    pred_matrix["person_dr", "fellow_dr"] <- 1
  }

  # Don't let fellow eye variables be predicted by anything
  pred_matrix[c("fellow_outcome_VA", "fellow_baseline_VA", "fellow_intermediate_VA", 
                "fellow_lapsed", "fellow_dr"), ] <- 0

  # Don't let fellow eye variables be their own predictors
  pred_matrix["fellow_outcome_VA", "fellow_outcome_VA"] <- 0
  pred_matrix["fellow_baseline_VA", "fellow_baseline_VA"] <- 0
  pred_matrix["fellow_intermediate_VA", "fellow_intermediate_VA"] <- 0
}

# Patient-level variables can predict eye-level outcomes (but not strongly)
if ("patient_worst_baseline" %in% colnames(pred_matrix)) {
  pred_matrix[, "patient_worst_baseline"] <- 1
  pred_matrix[, "patient_best_baseline"] <- 1
  pred_matrix[, "patient_ever_lapsed"] <- 1

  # But don't predict these variables
  pred_matrix["patient_worst_baseline", ] <- 0
  pred_matrix["patient_best_baseline", ] <- 0
  pred_matrix["patient_ever_lapsed", ] <- 0
}

# n_eyes and has_fellow are just indicators, not predictors
if ("n_eyes" %in% colnames(pred_matrix)) {
  pred_matrix[, "n_eyes"] <- 0
  pred_matrix["n_eyes", ] <- 0
  pred_matrix[, "has_fellow"] <- 0
  pred_matrix["has_fellow", ] <- 0
}

@@

```{r mice-run}
# Initialize logging
log_file <- init_log("multiple_imputation")
log_message(paste0("Analysis mode: ", analysis_mode))
log_message(paste0("Number of imputations: ", n_imputations))
log_message(paste0("Number of iterations per imputation: ", n_iterations))
log_message(paste0("Number of observations: ", nrow(df_mice)))
log_message(paste0("Variables with missing data: ", sum(colSums(is.na(df_mice)) > 0)))

log_message("Starting MICE imputation process...")

# Expected warnings about collinearity:
# 1. e_mrn_deidentified: Constant (patient ID) - expected, not used as predictor
# 2. outcome_VA_logMAR & fellow_outcome_VA: Identical missingness patterns (both NA for excluded cohort)
#    This is expected and OK - they're measuring different eyes but missing for same reason
# 3. Variables removed to avoid true collinearity: has_fellow, patient_ever_lapsed, fellow_lapsed
#    These were redundant with other variables in the model

imp_start <- Sys.time()

@@

  # Create summary table for cataract surgery proportions
  if(is.matrix(cat_props)) {
    cat_summary <- data.frame(
      Level = rownames(cat_props),
      Mean_Proportion = apply(cat_props, 1, mean) |> round(3),
      SD = apply(cat_props, 1, sd) |> round(3)
    )
  } else {
    cat_summary <- data.frame(
      Level = "Cataract Surgery",
      Mean_Proportion = round(mean(cat_props), 3),
      SD = round(sd(cat_props), 3)
    )
  }
  kable(cat_summary, caption = "Cataract surgery proportions across imputations")
}

# Fellow Eye Correlation Check
# For patients with both eyes, check correlation between eyes
if (sum(df_mice$n_eyes == 2, na.rm = TRUE) > 0) {

  # Extract first imputed dataset
  imp_data_1 <- complete(mice_results, 1)

  # Get bilateral patients
  bilateral_patients <- imp_data_1 |>
    filter(n_eyes == 2) |>
    arrange(e_mrn_deidentified)  # Sort by patient ID

  # Check correlation for baseline VA
  # Use slice_head and slice_tail to get first and second eye for each patient
  baseline_pairs <- bilateral_patients |>
    group_by(e_mrn_deidentified) |>
    summarise(
      eye1_baseline = first(baseline_VA_logMAR),
      eye2_baseline = last(baseline_VA_logMAR),
      .groups = "drop"
    ) |>
    filter(!is.na(eye1_baseline) & !is.na(eye2_baseline))

  baseline_cor <- cor(baseline_pairs$eye1_baseline, baseline_pairs$eye2_baseline)

  # Check correlation for outcome VA 
  outcome_pairs <- bilateral_patients |>
    group_by(e_mrn_deidentified) |>
    summarise(
      eye1_outcome = first(outcome_VA_logMAR),
      eye2_outcome = last(outcome_VA_logMAR),
      .groups = "drop"
    ) |>
    filter(!is.na(eye1_outcome) & !is.na(eye2_outcome))

  outcome_cor <- cor(outcome_pairs$eye1_outcome, outcome_pairs$eye2_outcome)

  # Summary table
  fellow_cor_summary <- data.frame(
    Measure = c("Baseline VA", "Outcome VA"),
    N_Pairs = c(nrow(baseline_pairs), nrow(outcome_pairs)),
    Correlation = c(round(baseline_cor, 3), round(outcome_cor, 3)),
    Expected_Range = c("0.7-0.9", "0.7-0.9"),
    Status = c(
      ifelse(baseline_cor >= 0.7 & baseline_cor <= 0.9, "✓ Within expected", "⚠ Outside expected"),
      ifelse(outcome_cor >= 0.7 & outcome_cor <= 0.9, "✓ Within expected", "⚠ Outside expected")
    )
  )

  kable(fellow_cor_summary, caption = "Fellow eye correlation in imputed data")

  # Visual check - scatter plot
  par(mfrow = c(1, 2))

  # Baseline VA correlation plot
  plot(baseline_pairs$eye1_baseline, baseline_pairs$eye2_baseline,
       xlab = "Eye 1 Baseline VA (logMAR)",
       ylab = "Eye 2 Baseline VA (logMAR)",
       main = paste0("Baseline VA Correlation (r = ", round(baseline_cor, 2), ")"),
       pch = 19, col = rgb(0, 0, 0, 0.5))
  abline(lm(eye2_baseline ~ eye1_baseline, data = baseline_pairs), col = "red", lwd = 2)

  # Outcome VA correlation plot
  plot(outcome_pairs$eye1_outcome, outcome_pairs$eye2_outcome,
       xlab = "Eye 1 Outcome VA (logMAR)",
       ylab = "Eye 2 Outcome VA (logMAR)",
       main = paste0("Outcome VA Correlation (r = ", round(outcome_cor, 2), ")"),
       pch = 19, col = rgb(0, 0, 0, 0.5))
  abline(lm(eye2_outcome ~ eye1_outcome, data = outcome_pairs), col = "red", lwd = 2)

  par(mfrow = c(1, 1))

  # Check correlation across all imputations
  all_cors <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    bilateral <- imp_data |> filter(n_eyes == 2)

    # Baseline correlation
    base_pairs <- bilateral |>
      group_by(e_mrn_deidentified) |>
      summarise(
        eye1 = first(baseline_VA_logMAR),
        eye2 = last(baseline_VA_logMAR),
        .groups = "drop"
      ) |>
      filter(!is.na(eye1) & !is.na(eye2))

    base_cor <- if(nrow(base_pairs) > 0) cor(base_pairs$eye1, base_pairs$eye2) else NA

    # Outcome correlation
    out_pairs <- bilateral |>
      group_by(e_mrn_deidentified) |>
      summarise(
        eye1 = first(outcome_VA_logMAR),
        eye2 = last(outcome_VA_logMAR),
        .groups = "drop"
      ) |>
      filter(!is.na(eye1) & !is.na(eye2))

    out_cor <- if(nrow(out_pairs) > 0) cor(out_pairs$eye1, out_pairs$eye2) else NA

    c(baseline_cor = base_cor, outcome_cor = out_cor)
  })

  # Summary of correlations across imputations
  cor_summary_all <- data.frame(
    Measure = c("Baseline VA", "Outcome VA"),
    Mean_Correlation = c(
      round(mean(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(mean(all_cors["outcome_cor",], na.rm = TRUE), 3)
    ),
    SD_Correlation = c(
      round(sd(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(sd(all_cors["outcome_cor",], na.rm = TRUE), 3)
    ),
    Min = c(
      round(min(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(min(all_cors["outcome_cor",], na.rm = TRUE), 3)
    ),
    Max = c(
      round(max(all_cors["baseline_cor",], na.rm = TRUE), 3),
      round(max(all_cors["outcome_cor",], na.rm = TRUE), 3)
    )
  )

  kable(cor_summary_all, caption = "Fellow eye correlations across all imputations")
}

# Clinical Plausibility Check for Imputed Values
log_message("Checking clinical plausibility of imputed values")

# Extract observed and imputed ranges for each variable
plausibility_checks <- list()

# Visual Acuity measures (logMAR scale)
# Clinical limits: -0.3 (exceptional vision) to 3.0 (hand motion/counting fingers)
va_vars <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR")

for (var in va_vars) {
  if (sum(is.na(df_mice[[var]])) > 0) {
    # Get observed range
    obs_range <- range(df_mice[[var]], na.rm = TRUE)

    # Get imputed ranges across all imputations
    imp_ranges <- sapply(1:n_imputations, function(i) {
      imp_data <- complete(mice_results, i)
      # Only look at previously missing values
      was_missing <- is.na(df_mice[[var]])
      if (sum(was_missing) > 0) {
        range(imp_data[[var]][was_missing])
      } else {
        c(NA, NA)
      }
    })

    plausibility_checks[[var]] <- data.frame(
      Variable = var,
      Observed_Min = round(obs_range[1], 3),
      Observed_Max = round(obs_range[2], 3),
      Imputed_Min = round(min(imp_ranges[1,], na.rm = TRUE), 3),
      Imputed_Max = round(max(imp_ranges[2,], na.rm = TRUE), 3),
      Clinical_Min = -0.3,
      Clinical_Max = 3.0,
      stringsAsFactors = FALSE
    )

    # Check if imputed values fall within observed range
    plausibility_checks[[var]]$Within_Observed <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Observed_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Observed_Max

    # Check if imputed values fall within clinical limits
    plausibility_checks[[var]]$Within_Clinical <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Clinical_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Clinical_Max
  }
}

# Comorbidity indices (non-negative integers)
comorbidity_vars <- c("CCI", "DCSI")

for (var in comorbidity_vars) {
  if (sum(is.na(df_mice[[var]])) > 0) {
    obs_range <- range(df_mice[[var]], na.rm = TRUE)

    imp_ranges <- sapply(1:n_imputations, function(i) {
      imp_data <- complete(mice_results, i)
      was_missing <- is.na(df_mice[[var]])
      if (sum(was_missing) > 0) {
        range(imp_data[[var]][was_missing])
      } else {
        c(NA, NA)
      }
    })

    plausibility_checks[[var]] <- data.frame(
      Variable = var,
      Observed_Min = round(obs_range[1], 1),
      Observed_Max = round(obs_range[2], 1),
      Imputed_Min = round(min(imp_ranges[1,], na.rm = TRUE), 1),
      Imputed_Max = round(max(imp_ranges[2,], na.rm = TRUE), 1),
      Clinical_Min = 0,
      Clinical_Max = ifelse(var == "CCI", 37, 13),  # CCI max 37, DCSI max 13
      stringsAsFactors = FALSE
    )

    plausibility_checks[[var]]$Within_Observed <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Observed_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Observed_Max

    plausibility_checks[[var]]$Within_Clinical <- 
      plausibility_checks[[var]]$Imputed_Min >= plausibility_checks[[var]]$Clinical_Min &
      plausibility_checks[[var]]$Imputed_Max <= plausibility_checks[[var]]$Clinical_Max
  }
}

# Combine all plausibility checks
if (length(plausibility_checks) > 0) {
  plausibility_summary <- do.call(rbind, plausibility_checks)
  rownames(plausibility_summary) <- NULL

  # Add status column
  plausibility_summary$Status <- ifelse(
    plausibility_summary$Within_Observed,
    "✓ Within observed range (PMM working correctly)",
    ifelse(
      plausibility_summary$Within_Clinical,
      "⚠ Outside observed but within clinical limits",
      "✗ Outside clinical limits"
    )
  )

  kable(plausibility_summary[, c("Variable", "Observed_Min", "Observed_Max", 
                                 "Imputed_Min", "Imputed_Max", "Status")],
        caption = "Clinical Plausibility of Imputed Values (PMM ensures values stay within observed range)")

  log_message(paste0("Plausibility check complete. All variables within observed range: ",
                    all(plausibility_summary$Within_Observed)))
}

# Check between-imputation variance
log_message("Checking between-imputation variance")

# Calculate variance across imputations for key outcomes
variance_checks <- sapply(va_vars, function(var) {
  if (sum(is.na(df_mice[[var]])) > 0) {
    # Get imputed values for each missing observation across all imputations
    was_missing <- which(is.na(df_mice[[var]]))

    if (length(was_missing) > 0) {
      imp_values <- sapply(1:n_imputations, function(i) {
        imp_data <- complete(mice_results, i)
        imp_data[[var]][was_missing]
      })

      # Calculate mean variance across observations
      if (is.matrix(imp_values)) {
        obs_variances <- apply(imp_values, 1, var)
        mean_var <- mean(obs_variances, na.rm = TRUE)
        min_var <- min(obs_variances, na.rm = TRUE)
        max_var <- max(obs_variances, na.rm = TRUE)
      } else {
        mean_var <- var(imp_values)
        min_var <- mean_var
        max_var <- mean_var
      }

      c(Mean_Variance = mean_var, Min_Variance = min_var, Max_Variance = max_var)
    } else {
      c(Mean_Variance = NA, Min_Variance = NA, Max_Variance = NA)
    }
  } else {
    c(Mean_Variance = NA, Min_Variance = NA, Max_Variance = NA)
  }
})

if (ncol(variance_checks) > 0) {
  variance_summary <- data.frame(
    Variable = colnames(variance_checks),
    Mean_Between_Imp_Variance = round(variance_checks["Mean_Variance",], 4),
    Min_Between_Imp_Variance = round(variance_checks["Min_Variance",], 4),
    Max_Between_Imp_Variance = round(variance_checks["Max_Variance",], 4),
    Status = ifelse(
      variance_checks["Mean_Variance",] > 0.001,
      "✓ Adequate variance",
      "⚠ Low variance - check predictor weights"
    )
  )

  kable(variance_summary, 
        caption = "Between-Imputation Variance (ensures uncertainty is preserved)")

  log_message(paste0("Variance check complete. Adequate variance in all variables: ",
                    all(variance_checks["Mean_Variance",] > 0.001, na.rm = TRUE)))
}
```

## MNAR Sensitivity Analysis: Delta-Shift Method

```{r mnar-sensitivity}
# Delta-shift sensitivity analysis for MNAR
# This shows how results would change if missing values were systematically different
# We do NOT apply these shifts to the main imputation - just show diagnostics

log_message("Performing MNAR sensitivity analysis (diagnostic only)")

# Define small symmetric deltas for outcome VA (in logMAR units)
# 0.1 logMAR = 1 line on eye chart
deltas <- c(-0.2, -0.1, 0, 0.1, 0.2)

# For each delta, show how the outcome distribution would shift
delta_results <- list()

for (delta in deltas) {
  # Take first imputed dataset as example
  imp_data_1 <- complete(mice_results, 1)

  # Identify which outcomes were imputed
  was_imputed <- is.na(df_mice$outcome_VA_logMAR)

  if (sum(was_imputed) > 0) {
    # Create shifted version (for diagnostic only)
    shifted_data <- imp_data_1
    shifted_data$outcome_VA_logMAR[was_imputed] <- 
      shifted_data$outcome_VA_logMAR[was_imputed] + delta

    # Calculate outcome statistics
    delta_results[[as.character(delta)]] <- data.frame(
      Delta = delta,
      Delta_Lines = round(delta / 0.1, 1),  # Convert to eye chart lines
      Mean_Observed = round(mean(imp_data_1$outcome_VA_logMAR[!was_imputed], na.rm = TRUE), 3),
      Mean_Imputed_Original = round(mean(imp_data_1$outcome_VA_logMAR[was_imputed], na.rm = TRUE), 3),
      Mean_Imputed_Shifted = round(mean(shifted_data$outcome_VA_logMAR[was_imputed], na.rm = TRUE), 3),
      Pct_VI_Observed = round(100 * mean(imp_data_1$outcome_VA_logMAR[!was_imputed] >= 1.0, na.rm = TRUE), 1),
      Pct_VI_Imputed_Original = round(100 * mean(imp_data_1$outcome_VA_logMAR[was_imputed] >= 1.0, na.rm = TRUE), 1),
      Pct_VI_Imputed_Shifted = round(100 * mean(shifted_data$outcome_VA_logMAR[was_imputed] >= 1.0, na.rm = TRUE), 1)
    )
  }
}

if (length(delta_results) > 0) {
  sensitivity_summary <- do.call(rbind, delta_results)
  rownames(sensitivity_summary) <- NULL

  # Add interpretation
  sensitivity_summary$Interpretation <- c(
    "Imputed have 2 lines better vision",
    "Imputed have 1 line better vision",
    "No systematic difference (MAR)",
    "Imputed have 1 line worse vision",
    "Imputed have 2 lines worse vision"
  )

  kable(sensitivity_summary[, c("Delta", "Delta_Lines", "Interpretation",
                                "Mean_Imputed_Original", "Mean_Imputed_Shifted",
                                "Pct_VI_Imputed_Original", "Pct_VI_Imputed_Shifted")],
        caption = "MNAR Sensitivity Analysis: Effect of systematic differences in missing outcomes",
        col.names = c("Delta (logMAR)", "Lines on chart", "Interpretation",
                     "Mean VA (Original)", "Mean VA (Shifted)",
                     "% VI (Original)", "% VI (Shifted)"))

  # Create visualization
  par(mfrow = c(1, 2))

  # Plot 1: Mean VA by delta
  plot(sensitivity_summary$Delta, sensitivity_summary$Mean_Imputed_Shifted,
       type = "b", pch = 19,
       xlab = "Delta adjustment (logMAR)",
       ylab = "Mean imputed outcome VA (logMAR)",
       main = "Sensitivity to MNAR assumption",
       ylim = range(c(sensitivity_summary$Mean_Observed,
                     sensitivity_summary$Mean_Imputed_Shifted)))
  abline(h = sensitivity_summary$Mean_Observed[1], col = "blue", lty = 2)
  abline(v = 0, col = "gray", lty = 3)
  legend("topleft", c("Shifted imputed", "Observed mean"), 
         col = c("black", "blue"), lty = c(1, 2), pch = c(19, NA))

  # Plot 2: % VI by delta
  plot(sensitivity_summary$Delta, sensitivity_summary$Pct_VI_Imputed_Shifted,
       type = "b", pch = 19,
       xlab = "Delta adjustment (logMAR)",
       ylab = "% with vision impairment",
       main = "VI prevalence under MNAR",
       ylim = range(c(sensitivity_summary$Pct_VI_Observed,
                     sensitivity_summary$Pct_VI_Imputed_Shifted)))
  abline(h = sensitivity_summary$Pct_VI_Observed[1], col = "blue", lty = 2)
  abline(v = 0, col = "gray", lty = 3)
  legend("topleft", c("Shifted imputed", "Observed prevalence"), 
         col = c("black", "blue"), lty = c(1, 2), pch = c(19, NA))

  par(mfrow = c(1, 1))

  log_message("MNAR sensitivity analysis complete (diagnostic only - main imputation unchanged)")
}
```
